---
title: "🌿 NDVI over Kealakekua Mountain Reserve with Sentinel-2 (R Tutorial)"
subtitle: "From Copernicus Browser to NDVI map in minutes"
author: "Ashley Holen"
format:
  html:
    theme: cosmo
    toc: true
    toc-title: "On this page"
    toc-depth: 3
execute:
  echo: true
  message: false
  warning: false
editor: visual
---

```{r}
library(here)
library(terra)
library(sf)
library(leaflet)
library(mapview)
```

## NDVI (Normalized Difference Vegetation Index)

NDVI is a simple, per-pixel indicator of “greenness” (photosynthetic activity).

### What it measures

Plants reflect more **near-infrared (NIR)** light and absorb **red** light. NDVI combines those two bands to highlight living vegetation.

### Formula

$$
\mathrm{NDVI} \;=\; \frac{\text{NIR} - \text{Red}}{\text{NIR} + \text{Red}}
$$

```{=html}
<!-- Plain-text fallback (if math is disabled):
NDVI = (NIR - Red) / (NIR + Red)
-->
```

### Typical value ranges

-   **0.6–0.9**: dense, healthy vegetation\
-   **0.2–0.5**: sparse/brown vegetation, shrubs/grass\
-   **0–0.2**: bare soil / urban\
-   **\< 0**: water, clouds, shadows, snow

### What data do we need to calculate NDVI

-   We can use the [Copernicus Open Access Hub](https://browser.dataspace.copernicus.eu/) to grab data form the satellite Sentinel 2

-   We specifically need Band 8 data for near-infared and Band 4 data for red

```{r}
# Paths to your .jp2 files
red_path2019 <- here("/Users/ashleyholen/Desktop/Geocomputation/DS422-Geocomputation/data/kona_data/new/2019/T04QHG_20190116T210921_B04_10m.jp2")
nir_path2019 <- here("/Users/ashleyholen/Desktop/Geocomputation/DS422-Geocomputation/data/kona_data/new/2019/T04QHG_20190116T210921_B08_10m.jp2")

red_path2020 <- here("/Users/ashleyholen/Desktop/Geocomputation/DS422-Geocomputation/data/kona_data/new/2020/T04QHG_20200225T210919_B04_10m.jp2")
nir_path2020 <- here("/Users/ashleyholen/Desktop/Geocomputation/DS422-Geocomputation/data/kona_data/new/2020/T04QHG_20200225T210919_B08_10m.jp2")

red_path2021 <- here("/Users/ashleyholen/Desktop/Geocomputation/DS422-Geocomputation/data/kona_data/new/2021/T04QHG_20211116T210919_B04_10m.jp2")
nir_path2021 <- here("/Users/ashleyholen/Desktop/Geocomputation/DS422-Geocomputation/data/kona_data/new/2021/T04QHG_20211116T210919_B08_10m.jp2")

red_path2022 <- here("/Users/ashleyholen/Desktop/Geocomputation/DS422-Geocomputation/data/kona_data/new/2022/T04QHG_20220301T210931_B04_10m.jp2")
nir_path2022 <- here("/Users/ashleyholen/Desktop/Geocomputation/DS422-Geocomputation/data/kona_data/new/2022/T04QHG_20220301T210931_B08_10m.jp2")

red_path2023 <- here("/Users/ashleyholen/Desktop/Geocomputation/DS422-Geocomputation/data/kona_data/new/2023/T04QHG_20230306T210921_B04_10m.jp2")
nir_path2023 <- here("/Users/ashleyholen/Desktop/Geocomputation/DS422-Geocomputation/data/kona_data/new/2023/T04QHG_20230306T210921_B08_10m.jp2")

red_path2024 <- here("/Users/ashleyholen/Desktop/Geocomputation/DS422-Geocomputation/data/kona_data/new/2024/T04QHG_20240214T210919_B04_10m.jp2")
nir_path2024 <- here("/Users/ashleyholen/Desktop/Geocomputation/DS422-Geocomputation/data/kona_data/new/2024/T04QHG_20240214T210919_B08_10m.jp2")
```

### These .jp2 files contain raster data

-   Raster data is a way to represent geography as a **grid of cells (pixels)**. Each cell has a single value (e.g., elevation, temperature, NDVI) and the whole grid is georeferenced so every cell aligns to a real-world location

-   The `terra` package helps us read raster data into R

```{r}
# Read bands directly
red2019 <- rast(red_path2019)
nir2019 <- rast(nir_path2019)

red2020 <- rast(red_path2020)
nir2020 <- rast(nir_path2020)

red2021 <- rast(red_path2021)
nir2021 <- rast(nir_path2021)

red2022 <- rast(red_path2022)
nir2022 <- rast(nir_path2022)

red2023 <- rast(red_path2023)
nir2023 <- rast(nir_path2023)

red2024 <- rast(red_path2024)
nir2024 <- rast(nir_path2024)
```

### Calculating NDVI

$$
\mathrm{NDVI} \;=\; \frac{\text{NIR} - \text{Red}}{\text{NIR} + \text{Red}}
$$

```{r}
ndvi2019 <- (nir2019 - red2019) / (nir2019 + red2019)

ndvi2020 <- (nir2020 - red2020) / (nir2020 + red2020)

ndvi2021 <- (nir2021 - red2021) / (nir2021 + red2021)

ndvi2022 <- (nir2022 - red2022) / (nir2022 + red2022)

ndvi2023 <- (nir2023 - red2023) / (nir2023 + red2023)

ndvi2024 <- (nir2024 - red2024) / (nir2024 + red2024)

```

### Plotting NDVI

```{r}
plot(ndvi2019, main = "NDVI 2019")
plot(ndvi2020, main = "NDVI 2020")
plot(ndvi2021, main = "NDVI 2021")
plot(ndvi2022, main = "NDVI 2022")
plot(ndvi2023, main = "NDVI 2023")
plot(ndvi2024, main = "NDVI 2024")
```

Looking at our results, it can be found that there seems to be an increase in NDVI around the reserve area over the time period of 2019-2025. It is key to note though, that the cloud coverage over the area does greatly impact the quality of the satellite imagery. Looking at the space that is not interfering with the cloud coverage, it can be seen that there is an increase in NDVI 2020-2023, but due to the image that we received in 2024, it seems as though NDVI decreased. Overall, it is difficult to form a conclusion when cloud coverage has such a large impact on the data available.
